---
title: "qc_control"
author: "Léonardo Panunzi"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---

```{r setup, echo=TRUE, cache=FALSE}
suppressWarnings(suppressMessages({
library(knitr)
library(rmdformats)
library(ComplexHeatmap)
library(gridExtra)
library(tidyr)
library(ggplot2)
library(reshape2)
library(readr)
library(dplyr)
library(openxlsx)
library(stringr)
library(viridis)
library(MASS)
library(nabor)
library(bigutilsr)
library(DESeq2)
library(biomaRt)
library(eulerr)
library(UpSetR)
}))
## Global options
options(max.print="1000")
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)

```

# Data loading

This analysis is based on several metrics from Picard output results (CollectRnaSeqMetrics).


```{r data_loading}
bam_met <- read.table("/datastage/silver/lysarc_bioinfo/RNAseq/data_input/bam_qc.tsv",header=TRUE, row.names = 1,sep="\t")
bam_met$EXONIC_READS_QUALIMAP <- gsub(".* /", "", bam_met$EXONIC_READS_QUALIMAP)
bam_met$INTRONIC_READS_QUALIMAP <- gsub(".* /", "", bam_met$INTRONIC_READS_QUALIMAP)
bam_met$INTERGENIC_READS_QUALIMAP <- gsub(".* /", "", bam_met$INTERGENIC_READS_QUALIMAP)
bam_met$NUMBER_OF_MAPPED_READS_QUALIMAP <- gsub(".* /", "", bam_met$NUMBER_OF_MAPPED_READS_QUALIMAP)
bam_met <- bam_met %>% mutate(across(everything(), as.numeric))
bam_mets <- scale(bam_met)
bam_mets <- bam_mets[,colSums(is.na(bam_mets))<nrow(bam_mets)]

```
# Scaled and clustered QC metrics:
### Pearson algorithm

```{r met_clustering, fig.width=11, fig.height=8}

p1 <- Heatmap(t(bam_mets), 
        name="scaled_value",
        col = circlize::colorRamp2(breaks = c(4, 0, -4), colors = c("red", "#EEEEEE", "blue")),
        show_column_names = F,
        show_row_names = T,
        clustering_distance_rows = "pearson",
        clustering_method_rows = "ward.D2")

print(p1)

```


# Data gathering according to QC metric

```{r dgqc, fig.width=11, fig.height=8}

pct_nub <- data.frame('val' = bam_met$PCT_USABLE_BASES)
rownames(pct_nub) <- rownames(bam_met)
pct_nub <- na.omit(pct_nub)
pct_nub_outliers <- boxplot(pct_nub$val, plot = FALSE)$out
pct_nub_out <- subset(pct_nub, pct_nub$val %in% pct_nub_outliers)
pct_nub_omin <- max(min(pct_nub$val), as.numeric(quantile(pct_nub$val, 0.25)) - (IQR(pct_nub$val)*1.5))
pct_nub_omax <- min(max(pct_nub$val), as.numeric(quantile(pct_nub$val, 0.75)) + (IQR(pct_nub$val)*1.5))
pct_nub <- pct_nub %>% mutate(status_ubs = ifelse(val >= pct_nub_omax, "outmax", ifelse(val <= pct_nub_omin, "outmin", "pass")))

pct_igb <- data.frame('val' = bam_met$INTERGENIC_BASES)
rownames(pct_igb) <- rownames(bam_met)
pct_igb <- na.omit(pct_igb)
pct_igb_outliers <- boxplot(pct_igb$val, plot = FALSE)$out
pct_igb_out <- subset(pct_igb, pct_igb$val %in% pct_igb_outliers)
pct_igb_omin <- max(min(pct_igb$val), as.numeric(quantile(pct_igb$val, 0.25)) - (IQR(pct_igb$val)*1.5))
pct_igb_omax <- min(max(pct_igb$val), as.numeric(quantile(pct_igb$val, 0.75)) + (IQR(pct_igb$val)*1.5))
pct_igb <- pct_igb %>% mutate(status_igb = ifelse(val >= pct_igb_omax, "outmax", ifelse(val <= pct_igb_omin, "outmin", "pass")))

pct_umr <- data.frame('val' = bam_met$UNIQUELY_MAPPED_READS_PCT_STAR)
rownames(pct_umr) <- rownames(bam_met)
pct_umr <- na.omit(pct_umr)
pct_umr_outliers <- boxplot(pct_umr$val, plot = FALSE)$out
pct_umr_out <- subset(pct_umr, pct_umr$val %in% pct_umr_outliers)
pct_umr_omin <- max(min(pct_umr$val), as.numeric(quantile(pct_umr$val, 0.25)) - (IQR(pct_umr$val)*1.5))
pct_umr_omax <- min(max(pct_umr$val), as.numeric(quantile(pct_umr$val, 0.75)) + (IQR(pct_umr$val)*1.5))
pct_umr <- pct_umr %>% mutate(status_umr = ifelse(val >= pct_umr_omax, "outmax", ifelse(val <= pct_umr_omin, "outmin", "pass")))

pct_fld <- data.frame('val' = bam_met$FAILED)
rownames(pct_fld) <- rownames(bam_met)
pct_fld <- na.omit(pct_fld)
pct_fld_outliers <- boxplot(pct_fld$val, plot = FALSE)$out
pct_fld_out <- subset(pct_fld, pct_fld$val %in% pct_fld_outliers)
pct_fld_omin <- max(min(pct_fld$val), as.numeric(quantile(pct_fld$val, 0.25)) - (IQR(pct_fld$val)*1.5))
pct_fld_omax <- min(max(pct_fld$val), as.numeric(quantile(pct_fld$val, 0.75)) + (IQR(pct_fld$val)*1.5))
pct_fld <- pct_fld %>% mutate(status_fld = ifelse(val >= pct_fld_omax, "outmax", ifelse(val <= pct_fld_omin, "outmin", "pass")))


load("/datastage/silver/lysarc_bioinfo/RNAseq/data_input/tximport.Rdata")
samples <- read.table("/datastage/silver/lysarc_bioinfo/RNAseq/data_input/sample_ids.tsv" ,header=TRUE, sep="\t")
samples_id <- read.table("/datastage/silver/lysarc_bioinfo/RNAseq/data_input/sample_ids_corresp.tsv" ,header=TRUE, sep="\t")
ddsTxi <- DESeqDataSetFromTximport(salmon_txi, colData = samples, design = ~1)
dds <- estimateSizeFactors(ddsTxi)
deseq_obj <- dds
normCounts <- log2(counts(deseq_obj, normalized=TRUE)+1)
KEEP.genes <- rowSums(edgeR::cpm(counts(deseq_obj)) >= 0.2) >= 10
normCountsAll <- normCounts
normCounts <- normCounts[KEEP.genes, samples$sample_id]
salmonCountsAll <- counts(dds, normalize=FALSE)

gtf <- read.delim(file=file("/datastage/silver/lysarc_bioinfo/RNAseq/data_input/gencode.v43.chr_patch_hapl_scaff.annotation.gtf"),
                    header=FALSE,
                    stringsAsFactors = FALSE,
                    comment.char = "#",
                    sep="\t")

gtf <- gtf[gtf$V3=="gene", ]

mySplit <- function(x, split="_", item=NULL, fixed=FALSE, perl=FALSE) {
  if(is.null(item)){
    sapply(x, function(y) unlist(strsplit(y, split=split, fixed=fixed, perl=perl)))
  } else{
    sapply(x, function(y) unlist(strsplit(y, split=split, fixed=fixed, perl=perl))[item])
  }
}

parseGTF <- function(gtf_row) {
    tmp <- mySplit(gsub("^ ","",mySplit(gtf_row[9], split=";")), split=" ")
    res <- tmp[2, ]
    names(res) <- tmp[1,]
    col_names <- c("gene_id", "gene_type", "gene_status", "gene_name", "level", "havana_gene") 
    # we try to get the value for the above columns (there are not present for all entries)
    res <- res[match(col_names, names(res))]
    names(res) <- col_names
    c(chr=gtf_row[[1]], start=gtf_row[[4]], end=gtf_row[[5]], strand=gtf_row[[7]], res, source=gtf_row[[2]])
}

gtf <- as.matrix(gtf)
annoGeneAll <- t(apply(gtf, 1, parseGTF))
annoGeneAll <- as.data.frame(annoGeneAll, stringsAsFactors=FALSE)
annoGeneAll$start <- as.numeric(annoGeneAll$start)
annoGeneAll$end <- as.numeric(annoGeneAll$end)
ensembl <- useEnsembl(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", version = 109)
biomartAnno <- getBM(attributes= c("ensembl_gene_id", "hgnc_symbol", "entrezgene_id", "gene_biotype", "description", "chromosome_name", "start_position", "end_position", "band", "strand"),
        filters=c("ensembl_gene_id"),
        values=list(gsub("\\..*","",annoGeneAll$gene_id)), mart=ensembl)
biomartAnno <- biomartAnno[order(biomartAnno$ensembl_gene_id, biomartAnno$entrezgene_id),]
annoGeneAll <- cbind(annoGeneAll, biomartAnno[match(gsub("\\..*","", annoGeneAll$gene_id), biomartAnno$ensembl_gene_id), c("ensembl_gene_id", "hgnc_symbol", "entrezgene_id",  "description","band", "gene_biotype")])
rownames(annoGeneAll) <- annoGeneAll$gene_id
prot_cod_genes <- annoGeneAll[annoGeneAll$gene_type == 'protein_coding' & annoGeneAll$gene_name %in% rownames(salmonCountsAll), ]
salmonCountsAll2 <- salmonCountsAll[rownames(salmonCountsAll) %in% prot_cod_genes$gene_name, ]  
rawLogCountsAll <- log2(salmonCountsAll2 + 1)
pca_raw <- prcomp(t(rawLogCountsAll))
annoSamples <- data.frame(sampleID=colnames(rawLogCountsAll))
annoSamples$percentNonZeroGenes <- colSums(rawLogCountsAll!=0)/nrow(rawLogCountsAll) *100
pct_gex <- data.frame('val' = annoSamples[[2]])
rownames(pct_gex) <- annoSamples[[1]]
pct_gex_outliers <- boxplot(pct_gex$val, plot = FALSE)$out
pct_gex_out <- subset(pct_gex, pct_gex$val %in% pct_gex_outliers)
pct_gex_omin <- max(min(pct_gex$val), as.numeric(quantile(pct_gex$val, 0.25)) - (IQR(pct_gex$val)*1.5))
pct_gex_omax <- min(max(pct_gex$val), as.numeric(quantile(pct_gex$val, 0.75)) + (IQR(pct_gex$val)*1.5))
annoSamples$gene_exp <- NA
annoSamples$gene_exp[annoSamples$percentNonZeroGenes<pct_gex_omin] <- "low"
annoSamples$gene_exp[annoSamples$percentNonZeroGenes>=pct_gex_omin] <- "intermediate"
annoSamples$gene_exp[annoSamples$percentNonZeroGenes>pct_gex_omax] <- "high"
col.gene_exp <- c(low="blue", intermediate="#EEEEEE", high="red")
corSamples <- cor(normCounts)
pct_gex <- pct_gex %>% mutate(status_gex = ifelse(val >= pct_gex_omax, "outmax", ifelse(val <= pct_gex_omin, "outmin", "pass")))
rownames(samples_id) <- samples_id$smp_id
pct_gexids <- merge(samples_id, pct_gex, by=0)
pct_gex <- data.frame(val=pct_gexids$val, status_gex=pct_gexids$status_gex)
rownames(pct_gex) <- pct_gexids[[2]]

qc_par <- merge(pct_nub, pct_igb, by=0)
rownames(qc_par) <- qc_par[[1]]
qc_par[[1]] <- NULL
qc_par$val.y <- NULL
qc_par$val.x <- NULL

qc_par <- merge(qc_par, pct_umr, by=0)
rownames(qc_par) <- qc_par[[1]]
qc_par[[1]] <- NULL
qc_par$val <- NULL

qc_par <- merge(qc_par, pct_fld, by=0)
rownames(qc_par) <- qc_par[[1]]
qc_par[[1]] <- NULL
qc_par$val <- NULL

qc_par <- merge(qc_par, pct_gex, by=0)
rownames(qc_par) <- qc_par[[1]]
gex <- data.frame(GENE_EXPRESSION=qc_par$val)
rownames(gex) <-rownames(qc_par)
qc_par[[1]] <- NULL
qc_par$val <- NULL

bam_mets_gex <- merge(bam_met, gex,  by=0)
rownames(bam_mets_gex) <- bam_mets_gex[[1]]
bam_mets_gex[[1]] <- NULL
bam_mets_gex <- scale(bam_mets_gex)
bam_mets_gex <- bam_mets_gex[,colSums(is.na(bam_mets_gex))<nrow(bam_mets_gex)]

qc_par <- merge(bam_mets_gex, qc_par,  by=0)
rownames(qc_par) <- qc_par[[1]]
qc_par[[1]] <- NULL

qc_par[qc_par == "outmin" | qc_par == "outmax"] <- "out"

qc_par_sort1 <- qc_par[order(-qc_par$PCT_USABLE_BASES), ]
qc_par_sort2 <- qc_par[order(-qc_par$INTERGENIC_BASES), ]
qc_par_sort3 <- qc_par[order(-qc_par$UNIQUELY_MAPPED_READS_PCT_STAR), ]
qc_par_sort4 <- qc_par[order(-qc_par$FAILED), ]
qc_par_sort5 <- qc_par[order(-qc_par$GENE_EXPRESSION), ]

```

# Outliers detection based on metric "PCT_USABLE_BASES"

```{r pct_nub, fig.width=11, fig.height=8}

hist(x = pct_nub$val, xlab='Proportion of Usable Bases', main = "", breaks = 150)
abline(v = median(pct_nub$val, na.rm = TRUE), col = "blue")

abline(v = pct_nub_omin, col = "red")
text(x=0.66, y=60, paste0('N = ', length(pct_nub_outliers[pct_nub_outliers<pct_nub_omin])), col='blue', cex=.95)

abline(v = pct_nub_omax, col = "red")
text(x=0.85, y=60, paste0('N = ', length(pct_nub_outliers[pct_nub_outliers>pct_nub_omax])), col='blue', cex=.95)

ha1 = HeatmapAnnotation(
    usable_bases = qc_par_sort1$status_ubs,
    intergenic_bases = qc_par_sort1$status_igb,
    unique_reads = qc_par_sort1$status_umr,
    failed = qc_par_sort1$status_fld,
    gene_expression = qc_par_sort1$status_gex,
    annotation_name_rot = 45,
    show_legend = F,
        col = list(
              usable_bases =  c("pass" = "orange", "out" = "grey30"),
              intergenic_bases =  c("pass" = "orange", "out" = "grey30"),
              unique_reads =  c("pass" = "orange", "out" = "grey30"),
              failed =  c("pass" = "orange", "out" = "grey30"),
              gene_expression =  c("pass" = "orange", "out" = "grey30")
    )
)

qc_par_sort1 <- qc_par_sort1[, -((ncol(qc_par_sort1)-4):ncol(qc_par_sort1))]

p1 <- Heatmap(t(qc_par_sort1), 
        name="scaled_value",
        col = circlize::colorRamp2(breaks = c(4, 0, -4), colors = c("red", "#EEEEEE", "blue")),
        top_annotation = ha1,
        cluster_columns = FALSE,
        show_column_names = F,
        show_row_names = T,
        clustering_distance_rows = "pearson",
        clustering_method_rows = "ward.D2")

print(p1)

```

# Outliers detection based on metric "INTERGENIC_BASES"

```{r pct_igb, fig.width=11, fig.height=8}
hist(x = pct_igb$val, xlab='Number of Intergenic Bases', main = "", breaks = 150)
abline(v = median(pct_igb$val, na.rm = TRUE), col = "blue")

abline(v = pct_igb_omin, col = "red")
text(x=5e+7, y=40, paste0('N = ', length(pct_igb_outliers[pct_igb_outliers<pct_igb_omin])), col='blue',  cex=.95)

abline(v = pct_igb_omax, col = "red")
text(x=6e+8, y=40, paste0('N = ', length(pct_igb_outliers[pct_igb_outliers>pct_igb_omax])), col='blue',  cex=.95)

ha2 = HeatmapAnnotation(
    usable_bases = qc_par_sort2$status_ubs,
    intergenic_bases = qc_par_sort2$status_igb,
    unique_reads = qc_par_sort2$status_umr,
    failed = qc_par_sort2$status_fld,
    gene_expression = qc_par_sort2$status_gex,
    annotation_name_rot = 45,
    show_legend = F,
        col = list(
              usable_bases =  c("pass" = "orange", "out" = "grey30"),
              intergenic_bases =  c("pass" = "orange", "out" = "grey30"),
              unique_reads =  c("pass" = "orange", "out" = "grey30"),
              failed =  c("pass" = "orange", "out" = "grey30"),
              gene_expression =  c("pass" = "orange", "out" = "grey30")
    )
)

qc_par_sort2 <- qc_par_sort2[, -((ncol(qc_par_sort2)-4):ncol(qc_par_sort2))]

p2 <- Heatmap(t(qc_par_sort2), 
        name="scaled_value",
        col = circlize::colorRamp2(breaks = c(4, 0, -4), colors = c("red", "#EEEEEE", "blue")),
        top_annotation = ha2,
        cluster_columns = FALSE,
        show_column_names = F,
        show_row_names = T,
        clustering_distance_rows = "pearson",
        clustering_method_rows = "ward.D2")

print(p2)

```

# Outliers detection based on metric "PCT_UNIQUELY_MAPPED_READS"

```{r pct_umr, fig.width=11, fig.height=8}

hist(x = pct_umr$val, xlab='Percentage of Uniquely Mapped Reads', main = "", breaks = 150)
abline(v = median(pct_umr$val, na.rm = TRUE), col = "blue")

abline(v = pct_umr_omin, col = "red")
text(x=44, y=25, paste0('N = ', length(pct_umr_outliers[pct_umr_outliers<pct_umr_omin])), col='blue', cex=.95)

abline(v = pct_umr_omax, col = "red")
text(x=85, y=25, paste0('N = ', length(pct_umr_outliers[pct_umr_outliers>pct_umr_omax])), col='blue', cex=.95)

ha3 = HeatmapAnnotation(
    usable_bases = qc_par_sort3$status_ubs,
    intergenic_bases = qc_par_sort3$status_igb,
    unique_reads = qc_par_sort3$status_umr,
    failed = qc_par_sort3$status_fld,
    gene_expression = qc_par_sort3$status_gex,
    annotation_name_rot = 45,
    show_legend = F,
        col = list(
              usable_bases =  c("pass" = "orange", "out" = "grey30"),
              intergenic_bases =  c("pass" = "orange", "out" = "grey30"),
              unique_reads =  c("pass" = "orange", "out" = "grey30"),
              failed =  c("pass" = "orange", "out" = "grey30"),
              gene_expression =  c("pass" = "orange", "out" = "grey30")
    )
)

qc_par_sort3 <- qc_par_sort3[, -((ncol(qc_par_sort3)-4):ncol(qc_par_sort3))]

p3 <- Heatmap(t(qc_par_sort3), 
        name="scaled_value",
        col = circlize::colorRamp2(breaks = c(4, 0, -4), colors = c("red", "#EEEEEE", "blue")),
        top_annotation = ha3,
        cluster_columns = FALSE,
        show_column_names = F,
        show_row_names = T,
        clustering_distance_rows = "pearson",
        clustering_method_rows = "ward.D2")

print(p3)

```


# Outliers detection based on metric "FAILED"

```{r pct_fld, fig.width=11, fig.height=8}

hist(x = pct_fld$val, xlab='Percentage of Uniquely Mapped Reads', main = "", breaks = 150)
abline(v = median(pct_fld$val, na.rm = TRUE), col = "blue")

abline(v = pct_fld_omin, col = "red")
text(x=0.007, y=100, paste0('N = ', length(pct_fld_outliers[pct_fld_outliers<pct_fld_omin])), col='blue', cex=.95)

abline(v = pct_fld_omax, col = "red")
text(x=0.032, y=100, paste0('N = ', length(pct_fld_outliers[pct_fld_outliers>pct_fld_omax])), col='blue', cex=.95)

ha4 = HeatmapAnnotation(
    usable_bases = qc_par_sort4$status_ubs,
    intergenic_bases = qc_par_sort4$status_igb,
    unique_reads = qc_par_sort4$status_umr,
    failed = qc_par_sort4$status_fld,
    gene_expression = qc_par_sort4$status_gex,
    annotation_name_rot = 45,
    show_legend = F,
        col = list(
              usable_bases =  c("pass" = "orange", "out" = "grey30"),
              intergenic_bases =  c("pass" = "orange", "out" = "grey30"),
              unique_reads =  c("pass" = "orange", "out" = "grey30"),
              failed =  c("pass" = "orange", "out" = "grey30"),
              gene_expression =  c("pass" = "orange", "out" = "grey30")
    )
)

qc_par_sort4 <- qc_par_sort4[, -((ncol(qc_par_sort4)-4):ncol(qc_par_sort4))]

p4 <- Heatmap(t(qc_par_sort4), 
        name="scaled_value",
        col = circlize::colorRamp2(breaks = c(4, 0, -4), colors = c("red", "#EEEEEE", "blue")),
        top_annotation = ha4,
        cluster_columns = FALSE,
        show_column_names = F,
        show_row_names = T,
        clustering_distance_rows = "pearson",
        clustering_method_rows = "ward.D2")

print(p4)

```


# Unusual proportion of gene expression : protein coding genes only

```{r expr_prop2, fig.width=11, fig.height=8}

hist(x = pct_gex$val, xlab='Percentage of Genes with Non-zero Expression', main = "", breaks = 150)
abline(v = median(pct_gex$val, na.rm = TRUE), col = "blue")

abline(v = pct_gex_omin, col = "red")
text(x=63, y=25, paste0('N = ', length(pct_gex_outliers[pct_gex_outliers<pct_gex_omin])), col='blue', cex=.95)

abline(v = pct_gex_omax, col = "red")
text(x=98, y=25, paste0('N = ', length(pct_gex_outliers[pct_gex_outliers>pct_gex_omax])), col='blue', cex=.95)


ha5 = HeatmapAnnotation(
    usable_bases = qc_par_sort5$status_ubs,
    intergenic_bases = qc_par_sort5$status_igb,
    unique_reads = qc_par_sort5$status_umr,
    failed = qc_par_sort5$status_fld,
    gene_expression = qc_par_sort5$status_gex,
    annotation_name_rot = 45,
    show_legend = F,
        col = list(
              usable_bases =  c("pass" = "orange", "out" = "grey30"),
              intergenic_bases =  c("pass" = "orange", "out" = "grey30"),
              unique_reads =  c("pass" = "orange", "out" = "grey30"),
              failed =  c("pass" = "orange", "out" = "grey30"),
              gene_expression =  c("pass" = "orange", "out" = "grey30")
    )
)

qc_par_sort5 <- qc_par_sort5[, -((ncol(qc_par_sort5)-4):ncol(qc_par_sort5))]

p5 <- Heatmap(t(qc_par_sort5), 
        name="scaled_value",
        col = circlize::colorRamp2(breaks = c(4, 0, -4), colors = c("red", "#EEEEEE", "blue")),
        top_annotation = ha5,
        cluster_columns = FALSE,
        show_column_names = F,
        show_row_names = T,
        clustering_distance_rows = "pearson",
        clustering_method_rows = "ward.D2")

print(p5)


p1 <- Heatmap(corSamples, show_row_names = FALSE, show_column_names = FALSE, name = "correlation", 
        top_annotation = columnAnnotation(df = annoSamples["gene_exp"], name="gene_exp", col=list(`gene_exp`=col.gene_exp) ))

print(p1)

jout_gex <-corSamples[pct_gex_out[[1]], pct_gex_out[[1]]]

p2 <- Heatmap(jout_gex,
              show_row_names = FALSE,
              show_column_names = FALSE,
              name = "correlation",
              )

print(p2)

gexpass <- corSamples[!rownames(corSamples) %in% rownames(pct_gex_out), ]

gex_pass <-corSamples[rownames(gexpass), rownames(gexpass)]
  
p3 <- Heatmap(gex_pass,
              show_row_names = FALSE,
              show_column_names = FALSE,
              name = "correlation",
              )

print(p3)

```


# UpSet plots

```{r upset, fig.width=11, fig.height=8}

qc_par_upset <- qc_par[, (ncol(qc_par)-4):ncol(qc_par)]
qc_par_upset[qc_par_upset == "out"] <- "1"
qc_par_upset[qc_par_upset == "pass"] <- "0"
qc_par_upset <- qc_par_upset %>% mutate(across(everything(), as.numeric))
qc_par_upset_out = qc_par_upset[rowSums(qc_par_upset[])>0,]

p1 <- upset(qc_par_upset, sets = c("status_ubs", "status_igb", "status_umr", "status_fld", "status_gex"), sets.bar.color = "#56B4E9",
order.by = "freq", empty.intersections = "on")

print(p1)

m = make_comb_mat(qc_par_upset)
m2 = make_comb_mat(qc_par_upset_out)

p3 <- UpSet(m2, comb_col = "#0000FF", bg_col = "#F0F0FF", bg_pt_col = "#CCCCFF")
print(p3)

ha <- HeatmapAnnotation("Intersection\nsize" = anno_barplot(comb_size(m2), border = FALSE, gp = gpar(fill = "black"), height = unit(3, "cm")), annotation_name_side = "left", annotation_name_rot = 0)

p4 <- UpSet(m2, top_annotation = upset_top_annotation(m2, gp = gpar(col = comb_degree(m2))))
print(p4)

p2 <- UpSet(m, comb_col = "#0000FF", bg_col = "#F0F0FF", bg_pt_col = "#CCCCFF")
print(p2)

```


# Unusual proportion of gene expression: all genes

```{r expr_prop, fig.width=11, fig.height=8}

load("/datastage/silver/lysarc_bioinfo/RNAseq/data_input/tximport.Rdata")
samples <- read.table("/datastage/silver/lysarc_bioinfo/RNAseq/data_input/sample_ids.tsv" ,header=TRUE, sep="\t")
samples_id <- read.table("/datastage/silver/lysarc_bioinfo/RNAseq/data_input/sample_ids_corresp.tsv" ,header=TRUE, sep="\t")

ddsTxi <- DESeqDataSetFromTximport(salmon_txi, colData = samples, design = ~1)
dds <- estimateSizeFactors(ddsTxi)
deseq_obj <- dds
normCounts <- log2(counts(deseq_obj, normalized=TRUE)+1)
KEEP.genes <- rowSums(edgeR::cpm(counts(deseq_obj)) >= 0.2) >= 10
normCountsAll <- normCounts
normCounts <- normCounts[KEEP.genes, samples$sample_id]
salmonCountsAll <- counts(dds, normalize=FALSE)

rawLogCountsAll <- log2(salmonCountsAll + 1)
pca_raw <- prcomp(t(rawLogCountsAll))

annoSamples <- data.frame(sampleID=colnames(rawLogCountsAll))
annoSamples$percentNonZeroGenes <- colSums(rawLogCountsAll!=0)/nrow(rawLogCountsAll) *100

pct_gex <- data.frame('val' = annoSamples[[2]])
rownames(pct_gex) <- annoSamples[[1]]

pct_gex_outliers <- boxplot(pct_gex$val, plot = FALSE)$out
pct_gex_out <- subset(pct_gex, pct_gex$val %in% pct_gex_outliers)
pct_gex_omin <- max(min(pct_gex$val), as.numeric(quantile(pct_gex$val, 0.25)) - (IQR(pct_gex$val)*1.5))
pct_gex_omax <- min(max(pct_gex$val), as.numeric(quantile(pct_gex$val, 0.75)) + (IQR(pct_gex$val)*1.5))

hist(x = pct_gex$val, xlab='Percentage of Genes with Non-zero Expression', main = "", breaks = 150)
abline(v = median(pct_gex$val, na.rm = TRUE), col = "blue")

abline(v = pct_gex_omin, col = "red")
text(x=25, y=25, paste0('N = ', length(pct_gex_outliers[pct_gex_outliers<pct_gex_omin])), col='blue', cex=.95)

abline(v = pct_gex_omax, col = "red")
text(x=45, y=25, paste0('N = ', length(pct_gex_outliers[pct_gex_outliers>pct_gex_omax])), col='blue', cex=.95)

pct_gex <- pct_gex %>% mutate(status_gex = ifelse(val >= pct_gex_omax, "outmax", ifelse(val <= pct_gex_omin, "outmin", "pass")))

rownames(samples_id) <- samples_id$smp_id
pct_gexids <- merge(samples_id, pct_gex, by=0)
pct_gex <- data.frame(val=pct_gexids$val, status_gex=pct_gexids$status_gex)
rownames(pct_gex) <- pct_gexids[[2]]


annoSamples$gene_exp <- NA
annoSamples$gene_exp[annoSamples$percentNonZeroGenes<pct_gex_omin] <- "low"
annoSamples$gene_exp[annoSamples$percentNonZeroGenes>=pct_gex_omin] <- "intermediate"
annoSamples$gene_exp[annoSamples$percentNonZeroGenes>pct_gex_omax] <- "high"
col.gene_exp <- c(low="blue", intermediate="#EEEEEE", high="red")
corSamples <- cor(normCounts)
corToMeanSample <- cor(normCounts, rowMeans(normCounts))

p1 <- Heatmap(corSamples, show_row_names = FALSE, show_column_names = FALSE, name = "correlation", 
        top_annotation = columnAnnotation(df = annoSamples["gene_exp"], name="gene_exp", col=list(`gene_exp`=col.gene_exp) ))

print(p1)

```
